Unitary evolution as a quasistatic process of infinitesimal collapses

### 2019-04-01 Articles/2019-ForQuantumOptics/article.tex
Adding Christine's comments on intro and conclusion

\begin{abstract}
Any state, to be properly defined, has to be an equilibrium of all processes happening at a faster timescale. As processes with equilibria are idempotent, they can be expressed as projections in a probability space. Assuming a quasistatic evolution, then, means the equilibria of the faster scale processes will evolve slowly, giving rise to continuous and unitary evolution. These ideas are enough to get the general form of the projection postulate and the time evolution of quantum mechanics. Violations of the Schroedinger equation, then, should be expected only for processes that cannot be considered quasistatic.
\end{abstract}

@@

In an effort to better understand why the frameworks of classical and quantum mechanics are what they are, we have worked on identifying a minimal set of physical assumptions from which both can be derived. This gave us a new perspective that is somewhat unusual among people who fork on fundamental issues in physics. The typical view is that the theories are so different that they require a radical different set of starting point, and that classical mechanics in particular should only be viewed as an approximation of the other. In this view, the similarities between the theories are due to one being an approximation of the other.

We believe a better view is to think of both classical and quantum mechanics as a specialization of a simpler more fundamental theory. This would deal with states and laws of evolution as abstract objects, formalizing what are the \emph{necessary} conditions for them to be operationally defined such that they can be experimentally studied. The other physical theories (classical and quantum mechanics, thermodynamics and so on) would specialize those concepts to better define what states we are studying and under what conditions. In this view, the similarities between the theories are due to the general concepts of the more fundamental theory. This different approach can help us pinpoint what elements are specific to what theory, clarify their realm of applicability and, hopefully, give us ideas where new physics may lie.

In this paper we start clarifying what are the fundamental relationships and, in particular, how they work in the context of quantum mechanics. The general idea is that for a state to be experimentally prepared, manipulated and measured, it needs to be ``stable'' for a maybe very short but certainly finite amount of time. That is, it must be a, possibly metastable, equilibrium of faster scale processes. But a process for which all outputs are equilibria is idempotent (i.e. gives the same output no matter how many is applied) and is therefore mathematically modeled by a projection operation on the space of statistical distributions. We identify quantum projections as these processes. If we want a deterministic law, then, it has to be a quasi-static evolution in terms of those fast scale processes. As the change has to be small, this leads to continuous evolution over the state space and unitary evolution in the space of statistical distributions. We identify unitary evolution as these processes.

The necessity of states to be stable for a certain minimum timescale applies to all physical theories, and a deterministic law, then, can only be valid \emph{above} that timescale within each theory. In other words, fast scale processes define the state space as their equilibria and the minimum scale $dt$ for the law of evolution. It will be instructive, then, to see how these concepts apply to classical mechanics and thermodynamics as well. This will foster the idea that looking at the foundations of all physical theories, and not on quantum mechanics in isolation, fosters a better general understanding of fundamental concepts.

In section 2 we will first take the basic framework of quantum mechanics and dress it in a slightly different language. The purpose is to make the interplay between pure states, statistical distributions of pure states, projections and unitary evolution more clear. In that context it will be easier to see how unitary evolution can be seen as a quasi-static process of infinitesimally close projections. In section 3 we will see how the same mathematical structure can be justified from more general physical consideration and how it necessarily applies to all physical theories, including classical mechanics in the thermodynamics.

@@

In this section we will assume as given the state space of non-relativistic quantum mechanics and the rules for projections. We will then recover unitary evolution (i.e. the Schroedinger equation) as the quasi-static deterministic evolution given by continuous projections. To do so, we will introduce a different mathematical notation which will make explicit the difference between pure states, statistical distributions and mixed states, to help dispel potential confusion. This notation will also prepare the stage for the next section, where we will see that the same mathematical framework applies to all scientific theories.

@@

If we accept this premise, then the natural thing to do is to write the conditions for a quasi-static deterministic evolution. That is, we assume that at each moment the system is in equilibrium and we change the state just a little so that each pure state is mapped to a single well-defined pure state. That is, if $\psi_{t}$ and $\psi_{t+dt}$ are two states infinitesimally close in time and $\mathcal{p}_{t+dt}$ is the process that between $[t, t+dt]$ transitions one into the other, then every time $\psi_{t}$ is the input of that process $\psi_{t+dt}$ is its output. Mathematically:
\begin{equation}
\begin{aligned}
\mathcal{p}_{t+dt}(\iota(\psi_{t}))&=\iota(\psi_{t+dt}) \\
\end{aligned}
\end{equation}
Applying the definitions we find:
\begin{equation}
\begin{aligned}
\mathcal{p}_{t+dt}(\iota(\psi_{t}))(\psi_{t+dt}) &=
\frac{\braket{\psi_{t} | \psi_{t+dt}} \braket{\psi_{t+dt}|\psi_{t}}}{\braket{\psi_{t}|\psi_{t}}} = 1 \\
\end{aligned}
\end{equation}
This requires the evolution to be unitary and we can use standard techniques from here, which will lead to the Schroedinger equation. In other words, \textbf{the Schroedinger equation is the quasi-static deterministic evolution of a process of continuous projections}. This approach, therefore, proceeds in the opposite way of others: instead of assuming the Schroedinger equation is always valid and deriving projections as a special case, we assume projections are valid approximations of faster scale processes and the Schroedinger equation is the special quasi-static deterministic case.

We can briefly see why this approach makes intuitive sense. The closer two states are in the Hilbert space, the greater the chance that one will transition to the other during a projection. Continuous evolution, then, means small deterministic changes. Unitary evolution, in particular, conserves entropy, which is also the hallmark of deterministic and reversible evolution. On the other hand, a projection can increase entropy, which is the hallmark of a transition that went out of equilibrium and is not reversible. Since any two unitary vectors can be connected by a unitary operation (i.e.~a rotation), it makes sense that all states are equilibria since they can all be connected by a deterministic and reversible process and that they all carry the same entropy. We also note that in quantum field theory, to calculate cross-sections and particle decays, one puts the initial states at minus infinity and the final states at plus infinity for processes that, in reality, take a tiny fraction of a second. This is reminiscent of a quasi-static approximation. Finally, while the approach presented here does not solve the so called ``measurement problem'', it does however give us an understanding of why we have the problem to begin with: our states are really just the equilibrium states of faster scale processes and those cannot, in general, be described by quasi-static ones.

@@

Every process that has equilibria is characterized by a time scale needed to reach equilibrium. That time will set the maximum time resolution we can use with that choice of states and processes. To go to faster time scales we inevitably need to change both states and processes. For example, as we change the volume of a gas, the region against the moving wall will be compressed sooner than the opposite region, but if we wait a short while we will again obtain a uniform pressure. If we want to study that transition, we may conceptually divide the system into small volumes and the pressure is replaced by a pressure field. Still, we need to assume each small volume to have enough molecules to be in equilibrium. We can go into more detail, and describe the distribution of position and velocities of the molecules. Yet, we need to assume that the molecules themselves are not undergoing chemical changes. We could study the atoms of the molecules, yet we would have to assume that the atoms are stable under nuclear fission. We could study the electrons and nucleons of the atoms, yet we would have to assume they are stable under weak nuclear forces. As we move to a finer and finer description, we have faster and faster scale processes that we assume do not change the basic constituents. At each level of the description what is assumed to be in equilibrium changes and what the laws describe changes as well. Yet, the overall structure remains the same: we have a timescale $dt$ such that within each interval $[t-dt, t]$ there is a process $\mathcal{p}_t$ such that the state $\mathcal{s}_t$ is an equilibrium of $\mathcal{p}_t$. The evolution $\mathcal{E}$, then, is quasi-static in terms of the processes in $\mathcal{P}$. That is, \textbf{all deterministic laws are quasi-static evolutions in terms of system-environment interaction processes}.

@@

In quantum mechanics we have elements of both. Like in classical mechanics we have a state space and a law of evolution, given by the choice of a Hamiltonian, that is deterministic and reversible. Like in thermodynamics, we have states that are equilibria of non-deterministic processes, of which we are not able to give a more fundamental description. Quantities are only well defined at equilibrium and the law of evolution is the quasi-static evolution where we assume the system changes slowly and we are always at equilibrium. What is interesting is that \textbf{in quantum mechanics what drives the deterministic evolution are the same processes that drive the equilibrium}.

@@

We have seen that quantum mechanics fits in a more general framework that applies to all other physical theories. States are always equilibria of some faster scale process, which in quantum mechanics are modeled with the projection. States are chosen to satisfy a law of evolution, which is necessarily a quasi-static process in terms of the faster scale processes, which in quantum mechanics is modeled by the Schroedinger equation. This general framework only uses standard physics concepts and puts no special role on the observer except the choice of what processes the system will be subjected to.

Violations of the Schroedinger equation, then, should be expected only for processes that cannot be considered quasi-static. If we were to implement a measurement process, then, in a way that the transfer of information between the system and the probe happens in a slow and controlled fashion, this would will not provide a violation to the Schroedinger equation, nor it would tell us that all projections are quasi-static processes. A better description of the faster scale processes given by the projections, then, will inevitable need a new state space as the standard quantum states only describes equilibria, and we should expect the faster scale processes to go out of equilibria.

To understand what is the right mathematical framework for those regimes, we are proceeding by first identifying the physical requirements for all mathematical structures already used in physics, starting from the most basic ones like topological spaces (upon which is based all of differential geometry) and $\sigma$-algebras (upon which is based all of measure theory and probability theory). This long term effort is centered mainly around the development of an open access book in which these issues can be explored in full detail and with a rigor more typical of mathematical disciplines. This rigor is what forces us to spell out the often hidden assumptions we have while constructing physical theories. By making explicit under which physically motivated circumstances time can be modeled by real numbers, states can be modeled by classical phase space or by quantum Hilbert spaces, we believe we will be in a better position to understand what happens when we can't.

@@

%The first main conclusion is that we should be able to find processes that are neither described by unitary evolution nor by the projection. These would correspond to processes for which the out-of-equilibrium dynamics becomes relevant so the quasi-static evolution is not a good approximation and neither is the instantaneous transition. Even if we do not have a theory for those regimes, we can, at least conceptually, discuss a couple of general strategies.

%For example, we can note that slow changing conditions are unlikely to reveal new dynamics. As long as the system evolves continuously, the quasi-static approximation will hold. It is also conceivable that, in some conditions, a process that may look like a projection end-to-end is actually well described by a slow transition. In those cases the quasi-static approximation evidently holds and no new phenomena would be observed. What is the right time scale, though, is not clear to us since ``slow'' in this setting is still ``very fast''.

>>>>

\begin{abstract}
Any state, to be properly defined, has to be an equilibrium of all processes happening at a faster timescale. As processes with equilibria are idempotent, they can be expressed as projections in a probability space. Assuming a quasistatic evolution, then, means the equilibria of the faster scale processes will evolve slowly, giving rise to continuous and unitary evolution. These ideas are sufficient to get the general form of the projection postulate and the time evolution of quantum mechanics. Violations of the Schroedinger equation, then, should be expected only for processes that cannot be considered quasistatic.
\end{abstract}

@@

In an effort to better understand why the frameworks of classical and quantum mechanics are what they are, we have worked on identifying a minimal set of physical assumptions from which both can be derived.[cite] This gave us a new perspective that is somewhat unusual among people who work on fundamental questions in physics. The typical view is that the theories are so different that they require a radically different set of starting points, and that classical mechanics in particular should only be viewed as an approximation of quantum mechanics. In this view, the similarities between the theories stem from one being an approximation of the other.

We believe a more encompassing view is to consider both classical and quantum mechanics as specializations of a simpler, more fundamental theory. This more fundamental theory would deal with states and laws of evolution as abstract objects, formalizing the \emph{necessary} conditions for them to be operationally defined such that they can be experimentally studied. The other physical theories (classical and quantum mechanics, thermodynamics and so on) would specialize those concepts to better define what states we are studying and under what conditions. In this view, the similarities between the theories are due to the general concepts of the more fundamental theory. This alternative approach can help us pinpoint the elements specific to each theory, clarify their realm of applicability and, hopefully, provide insights into where new physics may lie.

In this paper we outline the fundamental elements of a generic model for a physical system, their relationships and, in particular, how they work in the context of quantum mechanics. The general idea is that for a state to be experimentally prepared, manipulated and measured, it needs to be ``stable'' for what may be a very short but certainly finite amount of time. That is, it must be a, possibly metastable, equilibrium of faster scale processes. But a process for which all outputs are equilibria is idempotent (i.e. gives the same output no matter how many times it is applied) and is therefore mathematically modeled by a projection operation on the space of statistical distributions. We identify quantum projections as these processes. If we want a deterministic law, then, it must be a quasistatic evolution in terms of those faster scale processes. As the change must be small, this leads to continuous evolution over the state space and unitary evolution in the space of statistical distributions. We identify unitary evolution as this deterministic law.

The necessity of states to be stable for a certain minimum timescale applies to all physical theories, and a deterministic law, then, can only be valid at \emph{coarser} time resolution within each theory. In other words, faster scale processes define the state space as their equilibria and the minimum scale $dt$ for the law of evolution. It will be instructive, then, to see how these concepts apply to classical mechanics and thermodynamics as well. This will illustrate how looking at the foundations of all physical theories, and not at quantum mechanics in isolation, fosters a better general understanding of fundamental concepts.

In section 2 we will first take the basic framework of quantum mechanics and dress it in a slightly different language. The purpose is to make the interplay between pure states, statistical distributions of pure states, projections and unitary evolution clearer. In that context it will be easier to see how unitary evolution can be considered a quasistatic process of infinitesimally close projections. In section 3 we will see how the same mathematical structure can be justified from more general physical considerations and how it necessarily applies to all physical theories, including classical mechanics and thermodynamics.

@@

In this section we will assume as given the state space of non-relativistic quantum mechanics and the rules for projections. We will then recover unitary evolution (i.e. the Schroedinger equation) as the quasistatic deterministic evolution given by continuous projections. To do so, we will introduce a different mathematical notation which will make explicit the difference between pure states, statistical distributions and mixed states, to help dispel potential confusion. This notation will also prepare the stage for the next section, where we will see that the same mathematical framework applies to all scientific theories.

@@

If we accept this premise, then the natural thing to do is to write the conditions for a quasistatic deterministic evolution. That is, we assume that at each moment the system is in equilibrium and we change the state just a little so that each pure state is mapped to a single well-defined pure state. That is, if $\psi_{t}$ and $\psi_{t+dt}$ are two states infinitesimally close in time and $\mathcal{p}_{t+dt}$ is the process that between $[t, t+dt]$ transitions one into the other, then every time $\psi_{t}$ is the input of that process $\psi_{t+dt}$ is its output. Mathematically:
\begin{equation}
\begin{aligned}
\mathcal{p}_{t+dt}(\iota(\psi_{t}))&=\iota(\psi_{t+dt}) \\
\end{aligned}
\end{equation}
Applying the definitions we find:
\begin{equation}
\begin{aligned}
\mathcal{p}_{t+dt}(\iota(\psi_{t}))(\psi_{t+dt}) &=
\frac{\braket{\psi_{t} | \psi_{t+dt}} \braket{\psi_{t+dt}|\psi_{t}}}{\braket{\psi_{t}|\psi_{t}}} = 1 \\
\end{aligned}
\end{equation}
This requires the evolution to be unitary and we can use standard techniques from here, which will lead to the Schroedinger equation. In other words, \textbf{the Schroedinger equation is the quasistatic deterministic evolution of a process of continuous projections}. This approach, therefore, proceeds in the opposite way of others: instead of assuming the Schroedinger equation is always valid and deriving projections as a special case, we assume projections are valid approximations of faster scale processes and the Schroedinger equation is the special quasistatic deterministic case.

We can briefly see why this approach makes intuitive sense. The closer two states are in the Hilbert space, the greater the chance that one will transition to the other during a projection. Continuous evolution, then, means small deterministic changes. Unitary evolution, in particular, conserves entropy, which is also the hallmark of deterministic and reversible evolution. On the other hand, a projection can increase entropy, which is the hallmark of a transition that went out of equilibrium and is not reversible. Since any two unitary vectors can be connected by a unitary operation (i.e.~a rotation), it makes sense that all states are equilibria since they can all be connected by a deterministic and reversible process and that they all carry the same entropy. We also note that in quantum field theory, to calculate cross-sections and particle decays, one puts the initial states at minus infinity and the final states at plus infinity for processes that, in reality, take a tiny fraction of a second. This is reminiscent of a quasistatic approximation. Finally, while the approach presented here does not solve the so called ``measurement problem'', it does however give us an understanding of why we have the problem to begin with: our states are really just the equilibrium states of faster scale processes and those cannot, in general, be described by quasistatic ones.

@@

Every process that has equilibria is characterized by a time scale needed to reach equilibrium. That time will set the maximum time resolution we can use with that choice of states and processes. To go to faster time scales we inevitably need to change both states and processes. For example, as we change the volume of a gas, the region against the moving wall will be compressed sooner than the opposite region, but if we wait a short while we will again obtain a uniform pressure. If we want to study that transition, we may conceptually divide the system into small volumes and the pressure is replaced by a pressure field. Still, we need to assume each small volume to have enough molecules to be in equilibrium. We can go into more detail, and describe the distribution of position and velocities of the molecules. Yet, we need to assume that the molecules themselves are not undergoing chemical changes. We could study the atoms of the molecules, yet we would have to assume that the atoms are stable under nuclear fission. We could study the electrons and nucleons of the atoms, yet we would have to assume they are stable under weak nuclear forces. As we move to a finer and finer description, we have faster and faster scale processes that we assume do not change the basic constituents. At each level of the description what is assumed to be in equilibrium changes and what the laws describe changes as well. Yet, the overall structure remains the same: we have a timescale $dt$ such that within each interval $[t-dt, t]$ there is a process $\mathcal{p}_t$ such that the state $\mathcal{s}_t$ is an equilibrium of $\mathcal{p}_t$. The evolution $\mathcal{E}$, then, is quasistatic in terms of the processes in $\mathcal{P}$. That is, \textbf{all deterministic laws are quasistatic evolutions in terms of system-environment interaction processes}.

@@

In quantum mechanics we have elements of both. Like in classical mechanics we have a state space and a law of evolution, given by the choice of a Hamiltonian, that is deterministic and reversible. Like in thermodynamics, we have states that are equilibria of non-deterministic processes, of which we are not able to give a more fundamental description. Quantities are only well defined at equilibrium and the law of evolution is the quasistatic evolution where we assume the system changes slowly and we are always at equilibrium. What is interesting is that \textbf{in quantum mechanics what drives the deterministic evolution are the same processes that drive the equilibrium}.

@@

We have seen that quantum mechanics fits within a more general framework that applies to all physical theories. States are always equilibria of faster scale processes, which in quantum mechanics are modeled with the projection. States are chosen to satisfy a deterministic law of evolution, necessarily describing a quasistatic process, which in quantum mechanics is modeled by the Schroedinger equation. This general framework only uses standard physics concepts and puts no special role on the observer except the choice of what processes the system will be subjected to.

Violations of the Schroedinger equation, then, should be expected only for processes that cannot be considered quasistatic. If we were to implement a measurement process, then, in a way that the transfer of information between the system and the probe happens in a slow and controlled fashion, this could not provide a violation of the Schroedinger equation, nor would it tell us that all projections are quasistatic processes. A better description of the faster scale processes given by the projections, then, will inevitably need a new state space as the standard quantum states only describe equilibria, and we should expect the faster scale processes to go out of equilibrium. Ultimately, a description at Plank time scale will fail to be continuous and quasistatic since a finite change in a finite time will not be divided into infinitesimal adiabatic changes over infinitely many time steps.

To understand the suitable mathematical framework for those regimes where unitary evolution may not hold, we are proceeding by first identifying the physical requirements for all mathematical structures already used in physics, starting from the most basic ones like topological spaces (upon which is based all of differential geometry) and $\sigma$-algebras (upon which is based all of measure theory and probability theory). This long term effort is centered mainly around the development of an open access book in which these issues can be explored in full detail and with a rigor more typical of mathematical disciplines. This rigor forces us to spell out the often hidden assumptions we have while constructing physical theories. By making explicit, for example, under which physically motivated circumstances time can be modeled by real numbers, or when states can be represented within Hilbert spaces, we believe we will be in a better position to understand what happens when those assumptions fail.

@@

%The first main conclusion is that we should be able to find processes that are neither described by unitary evolution nor by the projection. These would correspond to processes for which the out-of-equilibrium dynamics becomes relevant so the quasistatic evolution is not a good approximation and neither is the instantaneous transition. Even if we do not have a theory for those regimes, we can, at least conceptually, discuss a couple of general strategies.

%For example, we can note that slow changing conditions are unlikely to reveal new dynamics. As long as the system evolves continuously, the quasistatic approximation will hold. It is also conceivable that, in some conditions, a process that may look like a projection end-to-end is actually well described by a slow transition. In those cases the quasistatic approximation evidently holds and no new phenomena would be observed. What is the right time scale, though, is not clear to us since ``slow'' in this setting is still ``very fast''.
